---
title: "DE_GO_RNA_seq"
author: "Johannes Schöneich"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(EnhancedVolcano)
library(tidyverse)
library(DESeq2)
library(limma)
library(glue)
library(clusterProfiler)
library(DOSE)
library(enrichplot)

harddrive <- "E:"

source(file.path(harddrive,"/Arbeit/helper_functions.R"))
source(file.path(harddrive,"/DE_analysis/DE_helperfunctions.R"))
```

```{r}
dds_fullsample <- readRDS(file.path(harddrive,
          "RNA_seq/RNAseq_uninfected/dds_LRT_RNAseq_uninfected.Rds"))

#condition to analyse: e.g. timepoint, condition
condition <- "timepoint"

analysis_folder <- "DE_analysis/DE_LRT_uninfected"

plots_folder <- file.path(harddrive,analysis_folder,"DE/plots")
charts_folder <- file.path(harddrive,analysis_folder,"DE/charts")

if (!dir.exists(analysis_folder)){
  dir.create(analysis_folder,recursive = TRUE)
  dir.create(plots_folder,recursive = TRUE)
  dir.create(charts_folder,recursive = TRUE)
}

# optional: read in the results.list if this analysis was already run 
# and you want to change plots
results.list <- readRDS(file.path(harddrive,analysis_folder,
                                  glue("DE_results.list_{condition}.Rds")))
```

## Volcano Plots DE genes

The contrast order is always +logFC then -logFC 
(first numerator (Zähler) then Denominator (Nenner))
condition treated vs untreated, tells you that the estimates are of the 
logarithmic fold change log2(treated/untreated).

The level given last is the base level for the comparison. The syntax is:
contrast <- c("condition", "level_to_compare", "base_level")
for example, if we observe a log2 fold change of -2 this would mean 
the gene expression is lower in the level of interest relative to the base level

```{r, warning=FALSE}
#every pairwise combination of timepoints
combination <- combn(levels(colData(dds_fullsample)[,condition]), 2)
list.names <- capture.output(apply(combination,2,function(x){cat(x,sep = "_",fill = TRUE)}))
list.names <- list.names[which(list.names!="NULL")]

if(!exists("results.list")){
#make an empty list with all combinations
results.list <- sapply(list.names,function(x) NULL)

for(i in 1:length(results.list)){
  print(i)
  #paste("working on", list.names[i])                                     
  results.list[[i]] <- results(dds_fullsample, contrast = c(condition, combination[1,i], #d1
                                                          combination[2,i]), #d5
  cooksCutoff = TRUE, independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH")

  results.list[[i]] <- lfcShrink(dds_fullsample, contrast = c(condition, combination[1,i], 
                                                            combination[2,i]), 
                       res = results.list[[i]],type = "ashr")
}

results.list <- lapply(results.list,function(x){x$symbol <- rownames(x);return(x)})
sapply(results.list,function(x){sum(is.na(x))}) #are there any NAs left?
results.list <- lapply(results.list,function(x){na.omit(x)})

#change logfc so that early is to the left.
results.list <- lapply(results.list,
                       function(res){
                         res$log2FoldChange <- -res$log2FoldChange
                         return(res)
                         }
                       )

saveRDS(results.list,file =
          file.path(analysis_folder,glue("DE_results.list_{condition}.Rds")
                    )
        )
}
```

```{r fig.height= 10, fig.width= 10}
pos_color <- "#E41A1C"
neg_color <- "#377EB8"

quantile_values <- sapply(results.list,function(x){
  quantile(x$log2FoldChange,c(0.001,0.999))
  }
)

xmin <- min(quantile_values)+1
xmax <- max(quantile_values)-1

p.cutoff = min(sapply(results.list,function(x){
  quantile(x[x$padj >0,]$padj,0.001)}))

for(i in 1:length(results.list)){
  res <- results.list[[i]]
  res.new <- as.data.frame(res)
  
  # extract the names from the results.list
  dayneg <- strsplit2(names(results.list[i]),split = "_")[1]
  daypos <- strsplit2(names(results.list[i]),split = "_")[2]
  
  # identify genes that pass the threshold
  up <- rownames(subset(res, log2FoldChange >= xmax))
  down <- rownames(subset(res, log2FoldChange <= xmin))
  p.up <- rownames(subset(res, padj <= p.cutoff))
  
  # change log2FCs in the object to be max or min cutoff 
  res.new$log2FoldChange <- ifelse(res.new$log2FoldChange >= xmax, xmax,
                                   ifelse(res.new$log2FoldChange <= xmin, xmin,
                                          res.new$log2FoldChange))
  # also change p.adj accordingly
  res.new$padj <- ifelse(res.new$padj <= p.cutoff, p.cutoff, res.new$padj)

  #top 20
  top20 <- res.new %>% top_n(20,log2FoldChange) %>% arrange(-log2FoldChange)
  
  #bot 20
  bot20 <- res.new %>% slice_min(log2FoldChange,n = 20) %>% arrange(-log2FoldChange)
  
  #Only those which are significant
  top20 <- subset(top20, subset = padj < 0.05 & log2FoldChange > 1) 
  bot20 <- subset(bot20, subset = padj < 0.05 & log2FoldChange < - 1)
    
  # custom shapes for the points
  customshape <- rep(19, nrow(res.new))
  names(customshape) <- rep('normal', nrow(res.new))
  customshape[which(rownames(res.new) %in% up)] <- -9658 #shape of the arrow facing right
  names(customshape)[which(rownames(res.new) %in% up)] <- '> LFC'
  customshape[which(rownames(res.new) %in% p.up)] <- 17 #shape of the arrow facing up
  names(customshape)[which(rownames(res.new) %in% p.up)] <- 'padj'
  customshape[which(rownames(res.new) %in% down)] <- -9668 #shape of the arrow facing left
  names(customshape)[which(rownames(res.new) %in% down)] <- '< LFC'
  
  # custom sizes for the points
  customsize <- rep(2.0, nrow(res.new))
  customsize [which(rownames(res.new) %in% up)] <- 4
  customsize [which(rownames(res.new) %in% p.up)] <- 4
  customsize [which(rownames(res.new) %in% down)] <- 4
  
  keyvals <- ifelse(
    res.new$log2FoldChange <= -1 & res.new$padj <= 0.05, neg_color,
    ifelse(res.new$log2FoldChange  >= 1 & res.new$padj <= 0.05, pos_color,
           'grey'))
  
  keyvals[is.na(keyvals)] <- 'grey'
  names(keyvals)[keyvals == neg_color] <- glue('up in {dayneg}')
  names(keyvals)[keyvals == 'grey'] <- 'ns'
  names(keyvals)[keyvals == pos_color] <- glue("up in {daypos}")

  # data.frame is needed for dplyr filter to work
  df <- as.data.frame(res[c("log2FoldChange","symbol","padj")]) %>% arrange(-log2FoldChange)

  up_genes <- nrow(df %>% filter(log2FoldChange >= 1 & padj <= 0.05) )
  down_genes <- nrow(df %>% filter(log2FoldChange <= -1 & padj <= 0.05))
  
  plt <- EnhancedVolcano(res.new,
                         lab = res.new$symbol,
                         x = 'log2FoldChange',
                         y = 'padj',
                         xlab = bquote(~Log[2]~'fold change'),
                         ylab = bquote(~-Log[10]~'adjusted p-value'),
                         selectLab = c(bot20$symbol,top20$symbol),
                         #drawConnectors = TRUE,
                         #arrowheads = FALSE,
                         max.overlaps = 0,
                         colCustom = keyvals,
                         pCutoff = 0.05,
                         FCcutoff = 1.0,
                         title = glue("DE genes {dayneg} vs. {daypos}"),
                         boxedLabels = FALSE,
                         colAlpha = 0.5,
                         labSize = 7,
                         axisLabSize = 35,
                         titleLabSize = 28,
                         subtitleLabSize = 24,
                         captionLabSize = 24,
                         legendLabSize = 24,
                         shapeCustom = customshape,
                         pointSize = customsize,
                         subtitle = glue("{dayneg}:{down_genes} {daypos}:{up_genes}"),
                         caption = paste0("total = ", nrow(res), " genes"),
                         #caption = NULL,
                         xlim = c(xmin,xmax)
  )
  
  save_ggplot_formats(
    plt = plt,
    base_plot_dir = plots_folder,
    plt_name=glue("Enhanced_volcano_{dayneg}_vs_{daypos}_quantile_cutoff_{condition}"),
    width = 7, height = 7
  )
  
df2 <- df %>% filter(log2FoldChange >= 1 & padj <= 0.05 | 
                       log2FoldChange <= -1 & padj <= 0.05)
write.csv(df2,file = file.path(charts_folder,
                                glue("DE_fullsamples_{dayneg}_vs_{daypos}_{condition}.csv")),
          row.names = FALSE)
}
```

# GO Terms with clusterProfiler

```{r}
# which ontology: choose between all, biological process (bp), molecular function (mf) 
# and cellular component (cc).
ont <- "bp"

#p-adjust cutoff. Adjustment method is benjamini-hochberg (false discovery rate)
pcutoff <- 0.05

go_plots_folder <- file.path(analysis_folder,"GO/plots")
go_charts_folder <- file.path(analysis_folder,"GO/charts")

if (!dir.exists(go_plots_folder)){
  dir.create(go_plots_folder,recursive = TRUE)
  dir.create(go_charts_folder,recursive = TRUE)
}
```

## GO analysis for upregulated Genes (positive values)

```{r}
for(list.nmbr in 1:length(results.list)){
  
GO_NAME <- paste(names(results.list)[list.nmbr],ont,pcutoff,sep = "_")
daypos <- strsplit2(names(results.list[list.nmbr]),split = "_")[2]

print(paste("calculating",daypos,"for",GO_NAME))

res <- results.list[[list.nmbr]]

#UP Regulation Table
sigsdf_UP <- res[res$log2FoldChange >= 1 & (res$padj <= 0.05),]

print(paste("number of genes",nrow(sigsdf_UP)))

UPgene <- (sigsdf_UP$symbol)
GO_upReg_results <- enrichGO(gene          = UPgene,
                             OrgDb         = "org.Mm.eg.db",
                             keyType       = "SYMBOL",
                             ont           = ont,
                             pAdjustMethod = "BH",
                             pvalueCutoff  = pcutoff,
                             readable      = TRUE)

  #arrange for p-value
  GO_upReg_results@result <- arrange(GO_upReg_results@result,p.adjust)

  GO_upReg_df <- as.data.frame(GO_upReg_results)

  if(nrow(GO_upReg_df) == 0){
    print(paste("No significant GO terms found for",GO_NAME))
    next
    }

  write.csv(GO_upReg_df,file = file.path(go_charts_folder,
                                     glue("GO_{daypos}_results_{GO_NAME}_{condition}.csv")),
            row.names = FALSE)
  if(0){
  if(nrow(GO_upReg_df) < 30){
    label_length = nrow(GO_upReg_df)
    } else{label_length = 30}
  
  labels <- str_wrap(rev(GO_upReg_df$Description[1:label_length]),width = 40)
  #dotplot colored by p-value and sorted reverse
  GO_upReg_Dotplot_pvalue <- dotplot.enrichResult2(GO_upReg_results,
                                                   showCategory = label_length, 
                                                   x = "p.adjust", title = GO_NAME,
                                                   color = "GeneRatio",font.size = 14) +
    scale_x_reverse() + #labels = scaleFUN
    scale_y_discrete(limits = rev(GO_upReg_df$Description[1:label_length]),
                     labels = labels) + 
    ggtitle(element_blank()) + 
    theme(legend.text = element_text(size = 16),legend.title = element_text(size=16),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  
  save_ggplot_formats(
    plt = GO_upReg_Dotplot_pvalue,
    create_plot_subdir = TRUE,
    base_plot_dir = go_plots_folder,
    plt_name = glue("GO_{daypos}_Dotplot_pvalue_{GO_NAME}_{condition}"),
    width = 8, height = 12
  )

  #netplot
  GO_upReg_Cnetplot <- cnetplot(GO_upReg_results, 
                                 showCategory = 10,max.overlaps = Inf) +
    ggtitle(glue("GO_{daypos}_Cnetplot_{GO_NAME}_{condition}")) +
    theme(plot.title = element_text(hjust =  0.5,face = "bold",size = 14))

  save_ggplot_formats(
    plt = GO_upReg_Cnetplot,
    create_plot_subdir = TRUE,
    base_plot_dir = go_plots_folder,
    plt_name=glue("GO_{daypos}_Cnetplot_{GO_NAME}_{condition}"),
    width = 12, height = 10
  )
  
  if(nrow(GO_upReg_df) < 5){
    print(paste("Not enough significant GO terms found for treeplot of",GO_NAME))
    next
  }
  #treeplot
  edox2 <- pairwise_termsim(GO_upReg_results)
  GO_upReg_enrichtreeplot <- treeplot(edox2, showCategory = label_length) + 
    ggtitle(glue("GO_{daypos}_enrichtreeplot_{GO_NAME}_{condition}")) + 
    theme(plot.title = element_text(hjust =  0.5,face = "bold",size = 14))
  
  save_ggplot_formats(
    plt = GO_upReg_enrichtreeplot,
    create_plot_subdir = TRUE,
    base_plot_dir = go_plots_folder,
    plt_name = glue("GO_{daypos}_enrichtreeplot_{GO_NAME}_{condition}"),
    width = 15, height = 10
  )}
}
```

## GO analysis for DOWNregulated Genes (negative values)

```{r}
for(list.nmbr in 1:length(results.list)){
  
  GO_NAME <- paste(names(results.list)[list.nmbr],ont,pcutoff,sep = "_")
  dayneg <- strsplit2(names(results.list[list.nmbr]),split = "_")[1]
 
  print(paste("calculating",dayneg,"for",GO_NAME))

  res <- results.list[[list.nmbr]]
  
  #DOWN Regulation Table
  sigsdf_DOWN <- res[(res$log2FoldChange <= -1) & (res$padj <= 0.05),]  
  print(paste("number of genes",nrow(sigsdf_DOWN)))
  
  if(nrow(sigsdf_DOWN) == 0){
    print(paste("No significant genes for",GO_NAME))
    next
  }
  
  DOWNgene <- (sigsdf_DOWN$symbol)
  
  GO_downReg_results <- enrichGO(gene = DOWNgene,
                                OrgDb = "org.Mm.eg.db",
                                keyType = "SYMBOL",
                                ont = ont,
                                pAdjustMethod = "BH",
                                pvalueCutoff  = pcutoff,
                                readable      = TRUE)

  #arrange for p-value
  GO_downReg_results@result<-arrange(GO_downReg_results@result,p.adjust)

  GO_downReg_df <- as.data.frame(GO_downReg_results)  
  
  if(nrow(GO_downReg_df) == 0){
    print(paste("No significant GO terms found for",GO_NAME))
    next
  }

  write.csv(GO_downReg_df,file = file.path(go_charts_folder,
              glue("GO_{dayneg}_results_{GO_NAME}_{condition}.csv"))
            ,row.names = FALSE)
  
  if(nrow(GO_downReg_df) < 30){
    label_length = nrow(GO_downReg_df)
    } else{label_length = 30}
  
  labels <- str_wrap(rev(GO_downReg_df$Description[1:label_length]),width = 40)
  
  #dotplot colored by p-value and sorted reverse
  GO_downReg_Dotplot_pvalue <- dotplot.enrichResult2(GO_downReg_results,
                                                     showCategory = label_length, 
                                                    x = "p.adjust", title = GO_NAME,
                          color = "GeneRatio",font.size = 14) + 
    scale_x_reverse() + #labels=scaleFUN
    scale_y_discrete(labels = labels, 
                     limits = rev(GO_downReg_df$Description[1:label_length])) +
    ggtitle(element_blank()) + 
    theme(legend.text = element_text(size = 16),legend.title = element_text(size = 16),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  
  
  save_ggplot_formats(
    plt = GO_downReg_Dotplot_pvalue,
    create_plot_subdir = TRUE,
    base_plot_dir = go_plots_folder,
    plt_name = glue("GO_{dayneg}_Dotplot_pvalue_{GO_NAME}_{condition}"),
    width = 8, height = 12
  )
 if(0){
  #netplot
  GO_downReg_Cnetplot <- cnetplot(GO_downReg_results,
                                 showCategory = 10,max.overlaps = Inf) +
    ggtitle(glue("GO_{dayneg}_Cnetplot_{GO_NAME}")) +
    theme(plot.title = element_text(hjust =  0.5,face = "bold",size = 14))

  save_ggplot_formats(
    plt = GO_downReg_Cnetplot,
    create_plot_subdir = TRUE,
    base_plot_dir = go_plots_folder,
    plt_name=glue("GO_{dayneg}_Cnetplot_{GO_NAME}_{condition}"),
    width = 12, height = 10
  )

  if(nrow(GO_downReg_df) < 5){
    print(paste("Not enough significant GO terms found for treeplot of",GO_NAME))
    next
  }
  #treeplot
  edox2 <- pairwise_termsim(GO_downReg_results)
  GO_downReg_enrichtreeplot <- treeplot(edox2) + 
    ggtitle(glue("GO_{dayneg}_enrichtreeplot_{GO_NAME}_{condition}")) +
    theme(plot.title = element_text(hjust =  0.5,face = "bold",size = 14))
  
  save_ggplot_formats(
    plt = GO_downReg_enrichtreeplot,
    create_plot_subdir = TRUE,
    base_plot_dir = go_plots_folder,
    plt_name = glue("GO_{dayneg}_enrichtreeplot_{GO_NAME}_{condition}"),
    width = 15, height = 10
  )
 }
}
```